How predictable is extinction?

### 2018-11-30 doc/intro.tex
adding more details to methods

Species global geographic range changes over time due to both local and global processes.

An old assumption in paleobiology was that species quickly obtained their maximum global geographic range and then maintained that range till their decline eventual extinction. Observation of species distribution records does not support this assumption. Instead, changes to a species' global geographic range is currently theorized to follow one of two general patterns: hat or hump PERFECT TIME FOR A FIGURE. Both are symmetric patterns but have different shapes. The hat pattern is characterized by an increase in geographic range for the first half a species' duration followed by a brief peak in geographic range, and the a decrease in geographic range for the second half of that species' duration. The hump pattern is instead characterized by a brief period of increasing species geographic range, an extended period of that species' duration at its maximum geographic range, and finally a quick decline and extinction of the species.

Measuring this pattern is extremely difficult for multiple reasons. Species are not observed continuously through time, species geographic ranges are based on what samples are avaliable, and species are harder to observe at the tails of their duration because they are rare (small geographic range, small population sizes).

Additionally, neither of these patterns are necessarily different from a Markovian process with a lower bound. A random walk beginning at 0 will most likely be unimodal with a period of increase followed by a period of decrease and then eventually reaching 0. PERFECT PLACE FOR A SIMULATION.

\end{document}

>>>>

Instead of modeling geographic range through time, we instead model extinction probability as a function of change in geographic range. This formulation means that, if there if geographic range trajectory is deterministic and not entirely Markovian, that past geographic range, specifically the change in geographic range, and not just current geographic range predicts the probability of extinction. For example, is extinction probability affected by a species has been decreasing in geographic range over time or does it just matter if that species has below average geographic range regardless of how it got there?

Rarely are we able to analyze long periods of geological time at fine resolutions -- below the 5-10 My scale. Due to substantial effort and the unique biology of the system, the microfossil provides us the unique opportunity to analyze ecological and evolutionary patterns at approximately million-year time scales. Typical ``exceptional'' fossil records tend to be of individual taxonomic groups and for rarely longer than 10 million years. The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores the occurrences are recorded from -- there is no equivalent fossil record. By analyzing patterns of extinction and global occurrence at fine temporal scales, we can better elucidate how well we can predict species extinction at human-relevant scales.

### 2018-11-30 doc/methods.tex
added more to methods and results...tried at least
adding more details to methods

Temperature data used as covariates in this analysis are based on Magnesium/Calcium isotope ratios sourced from Cramer et al CITATION. These are considered a more accurate estimate of global temperature than the standard Oxygen isotope-based estimates because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g. meteoric water). This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, Cramer et al., provides temperature estimates for every 0.1 My from 0 to 63 Mya. We binned these estimates into 1 My intervals as we did with the fossil occurrences. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval (Fig. \ref{fig:temp_curve}). Temperature was then transformed and standardized the in the same manner as geographic range (above).

@@

An implementation of our full model in \begin{texttt}rstanarm\end{texttt}, given a data.frame of all necessary data in a data.frame called ``data'', is coded as
\begin{verbatim}
  stan_glmer(formula = event ~ range + range_diff + temp + temp_lag + 
                       (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
                       (1 | age/phylum), 
             data = data, 
             family = 'binomial',
             prior = normal(c(-1, 0, 0, 0), rep(1, 4), autoscale = FALSE), 
             prior_intercept = normal(0, 10, autoscale = FALSE), 
             prior_aux = cauchy(0, 3, autoscale = FALSE), 
             chains = 4,
             thin = 4,
             adapt_delta = 0.999999).
\end{verbatim}

Similarly, our full model can be implemented in \begin{texttt}brms\end{texttt} as
\begin{verbatim}
priors <- c(set_prior('normal(0, 10)', class= 'Intercept'),
            set_prior('normal(0, 1)', class = 'b'),
            set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
            set_prior('cauchy(0, 3)', class = 'sd'),
            set_prior('lkj(1)', class = 'cor'))
brmfit <- brm(formula = bf(event ~ range + range_diff + temp + temp_lag +
                           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
                           (1 | age/phylum))
              data = data, 
              family = bernoulli(), 
              prior = priors,
              chains = 4, 
              thin = 4,
              control = list(adapt_delta = 0.999999).
\end{verbatim}

@@

The four variant models were compared using both WAIC and LOOIC which are estimates of a model's expected out-of-sample performance; these measures are fully Bayesian, taking into account the entire estimated joint posterior CITATION. WAIC and LOOIC are interpreted similarly to AIC, with lower values indicating greater expected out-of-sample performance. Additionally, the Bayesian nature of WAIC and LOOIC mean that they are calculated with standard errors allowing for a more nuanced understanding the actual amount of difference between the models. We selected the model with the lowest WAIC and LOOIC, which happened to be are most parameter rich model.

Model adequacy was measured using the area under the receiver operating characteristic curve (AUC). This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to only true positive rate measured by accuracy CITATION. This value ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance. AUC was calculated for the dataset as a whole, and for each of time bins. This measure was used in two contexts: measuring in-sample and out-of-sample predictive performance. ``In-sample'' means we are estimating how well our model predicts our observed data given that the model was fit to the entire dataset; this is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data. ``Out-of-sample'' is defined below.

We are also interested in understanding how well our model predicts species extinction given new, future data (out-of-sample data). To do this, we estimated average out-of-sample predictive error using 5-fold time-series cross-validation. For time-series data, the folds (data partitions) are approximately equal segments of time. The model is fit to the first fold and the posterior estimates are used to predict the states of the observations in the second fold, then the model is fit to the first and second fold and the posterior states are used to estimate the states from the third fold, and so on with increasingly large numbers of folds used for fitting a model to predict the states from the subsequent fold. With 63 time points, each of the five folds represents approximately 13 time points. Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.

>>>>

Temperature data used as covariates in this analysis are based on Magnesium/Calcium isotope ratios sourced from Cramer et al CITATION. These elemental ratios are considered more accurate estimates of past global temperature when compared to the frequently used Oxygen isotope-based estimates; this is because Mg/Ca based estimates are not effected by ice-volume and fresh-water input (e.g. meteoric water) which can alter Oxygen isotope ratios without reflecting changes to the climate itself. This property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, Cramer et al., provides temperature estimates for every 0.1 My from 0 to 63 Mya. We binned these estimates into 1 My intervals as we did with the fossil occurrences. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval (Fig. \ref{fig:temp_curve}). Temperature was then transformed and standardized the in the same manner as geographic range (above).

@@

An implementation of our full model in \begin{texttt}rstanarm\end{texttt}, given a data.frame of all necessary data in a data.frame called ``data'', is coded as:
\begin{verbatim}
  stan_glmer(formula = event ~ range + range_diff + temp + temp_lag + 
                       (1 + range + range_diff + temp + temp_lag | mybin/phylum) + 
                       (1 | age/phylum), 
             data = data, 
             family = 'binomial',
             prior = normal(c(-1, 0, 0, 0), rep(1, 4), autoscale = FALSE), 
             prior_intercept = normal(0, 10, autoscale = FALSE), 
             prior_aux = cauchy(0, 3, autoscale = FALSE), 
             chains = 4,
             thin = 4,
             adapt_delta = 0.999999)
\end{verbatim}

Similarly, our full model can be implemented in \begin{texttt}brms\end{texttt} CITATION as:
\begin{verbatim}
priors <- c(set_prior('normal(0, 10)', class= 'Intercept'),
            set_prior('normal(0, 1)', class = 'b'),
            set_prior('normal(-1, 1)', class = 'b', coef = 'range'),
            set_prior('cauchy(0, 3)', class = 'sd'),
            set_prior('lkj(1)', class = 'cor'))
brmfit <- brm(formula = bf(event ~ range + range_diff + temp + temp_lag +
                           (1 + range + range_diff + temp + temp_lag | mybin/phylum) +
                           (1 | age/phylum))
              data = data, 
              family = bernoulli(), 
              prior = priors,
              chains = 4, 
              thin = 4,
              control = list(adapt_delta = 0.999999)
\end{verbatim}

@@

We are interested in model adequacy and performance into two contexts: in-sample and out-of-sample predictive performance. ``In-sample'' means we are estimating how well our model predicts our observed data given that the model was fit to the entire dataset; this is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data. ``Out-of-sample'' is defined below.

Relative and absolute model adequacy of the four variant models was compared using the area under the receiver operating characteristic curve (AUC). This measure is commonly used in classification problems as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers the count of true positives. AUC ranges between 0.5 and 1, with 0.5 indicating no improvement in performance from random and 1 indicating perfect performance.

The differences in in-sample predictive performance between the models was visualized in multiple ways: whole data set by model, taxonomic group by model, model performance over time, and model performance by taxonomic groups over time. These comparisons demonstrate the relative and absolute adequacy of the models in describing the dataset they were fit to.

We are particularly interested in understanding how well our model predicts species extinction given new, future data (out-of-sample data). To do this, we estimated average out-of-sample predictive error using 5-fold time-series cross-validation. For time-series data, the folds (data partitions) are approximately equal segments of time. The model is fit to the first fold and the posterior estimates are used to predict the states of the observations in the second fold, then the model is fit to the first and second fold and the posterior states are used to estimate the states from the third fold, and so on with increasingly large numbers of folds used for fitting a model to predict the states from the subsequent fold. With 63 time points, each of the five folds represents approximately 13 time points. Keep in mind, however, that each time point corresponds to many (100-1000) individual observations.
